The Bot on Speaking Terms: The Effects of Conversation Architecture on Perceptions of Conversational Agents

### 2023-02-18 paper.tex
reworked introduction based on AK comments

Conversational agents (CAs), such as Amazon Alexa or Siri, are designed to interact with humans using natural language through voice or text modalities. %Adoption of CAs such as Amazon Alexa or Siri (Apple) is rapidly growing in the market, with half of Internet users in United States owning one or more smart speaker devices \cite{2022comscore}. These conversational agents are designed to be ubiquitous, easily accessible through smartphones, tablets, laptops and smart devices. Since these agents interact with users using natural language, there are nuances in their communications affected by conversation architecture, which is what they say and how they say it.
%they solicit social responses from users, encouraging users to attribute lifelike qualities, a.k.a. anthropomorphized perceptions to CAs \cite{eyssel2012if}.
%Anthropomorphism is defined as the attribution of traits that we typically associate with being distinctly human to nonhuman entities \cite{waytz2010sees}. For example, users tend to personify conversational agents like Amazon Alexa by using person pronouns instead of object pronouns in online reviews \cite{purington2017alexa}.  Research shows that anthropomorphic design is beneficial to establishing and maintaining trust between users and conversational agents \cite{seeger2021chatbots}. Also, speakers tend to align their lexical choices with conversational agents, similar to human-human conversations \cite{cowan2015does}. Anthropomorphism has been shown to ease user interactions, e.g. making agents more approachable, engaging, and trustworthy when they exhibit human characteristics \cite{qiu2009evaluating}. These processes are of particular interest for human-agent communication because they form the basis for user expectations and predispositions regarding agents' behavior, reliability of the information, etc. \cite{kuzminykh2020genie}.
Previous research has shown that there is a connection between conversation architecture and the perception of agents. Studies have found that conversational agents are perceived as more socially present and emotionally intelligent if they use affect language such as sentiment-adaptive response based on user's utterances, resulting in higher user satisfaction \cite{diederich2019emulating}\cite{yang2017perceived}. Also, CAs that mimic user's utterances through lexical alignment reduce users' cognitive workload, resulting in higher engagement with the interaction \cite{spillner2021talk}. Non-linguistic cues such as expressive prosody contribute to higher perceived intimacy with the user, as well as higher perceived enjoyment and ease of use \cite{kim2020can}. Although research on how conversation architecture affects agents' perceptions already exists, there is a lack of understanding on the coverage of relationships explored in literature, as well as synthesizing these findings into a generalized framework. One of the possible issues is the lack of clear and consistent definition on the aspects of user's perceptions of agent and conversation architectural elements.
%However, conversational architecture design may lead to negative consequences, such as an agent with an extroverted personality may be seen as less trustworthy, as users are irritated by the system's overly familiar attitude \cite{andrews2012system}.

Related to the perception aspects of agents, there have been recent efforts to systematically analyze user experience criteria used to evaluate conversational agents. Some review papers have created a list of dimensions used to assess conversational agents, which include the perceptions of agents (e.g. empathy, humanness) alongside other assessments (e.g. physiological data) \cite{clark2019state}\cite{finch2020towards}.
%In Clark et al's review for the state of speech interfaces, the authors created a list the different concepts used to assess disembodied voice-based agents such as user attitudes and task performance \cite{clark2019state}. 
Other papers categorized CA assessments into themes. For example, Rapp et al. \cite{rapp2021human} generated themes related to how users' experience with chatbots such as experiencing the chatbot including assessments on expectation, perception, satisfaction etc. Most recently, Zheng et al's literature review on the UX research of CAs \cite{zheng2022ux} categorized the metrics based on the type of method used, including technical measures like system error rate and perception measures like perceived social presence. However, existing studies use different terminologies and categories across aspects of perception. As an illustration, it is unclear if the theme of experiencing the chatbot from Rapp et al. \cite{rapp2021human} contains similar aspects to the perception towards CA in Zheng et al \cite{zheng2022ux}'s analysis.

For conversation architecture elements, Feine et al \cite{feine2019taxonomy} created foundational knowledge by providing a taxonomy of social cues, which is defined as "a cue of a CA that triggers a social reaction of the user towards the CA." Elements of conversation architecture are captured in the taxonomy, alongside other social cues such as the embodied agents' appearance and movements. Clark et al \cite{clark2019state} had also captured conversation architecture elements under the research topic of system speech production, with similar categorizations across content, style, and other topics. While these papers captured conversation architecture elements as part of their broader scope of research, there exist the need for standardized definitions specifically for conversation architecture that reflects its unique characteristics.

%there is existing research to systematically analyze them, like Feine et al \cite{feine2019taxonomy}'s overall framework provides consistent categories for social cues. However, it is unknown which conversation architecture elements have impacts on the perceptions of agents. Lastly, while there are various studies investigating the effect of  conversational architecture elements on the anthropomorphized perception of agents, we did not find any systematic review and analysis to generalize the findings into a usable framework.

In this paper, we are addressing these research gaps by analyzing how to design conversation architecture in order to orchestrate the perceptions of conversational agents. First, we want to understand \textbf{(RQ1) what perception aspects of agents are being explored in relation to the effects of conversation architecture elements}. Various elements of conversation architecture can be used in a CA's utterances such as humour, emoticons, or lexical alignment. This leads to our second research question: \textbf{(RQ2) what elements of conversation architecture are relevant to the perceptions of agents?} Lastly, we want to understand the impact of conversation architecture on perceptions of agent, which leads to our third research question: \textbf{(RQ3) what are the relationships between the specifics of conversation architecture elements and the perceptions of CAs?}

To answer these questions, we performed a comprehensive review of existing literature from 2010 to 2022 and synthesized findings on the impact of conversation architecture elements on the perceptions of CAs based on relevant papers (n=57). We collected perceptions of agents that are affected by conversation architecture,  grouped the aspects into four categories and eleven sub-categories (Table \ref{tab:perceptions}). We also analyzed conversation architecture related to perception of agents and grouped them into three categories and ten elements (Table \ref{tab:cues}). Lastly, we coded the explored connections between conversation architecture and perceptions and demonstrated the relationship into a synthesized framework (Figure \ref{fig:heatmap-impact}). Through our analysis, we also found that there are inconsistencies in the assessment of perceptions across studies, as well as influencing factors other than conversation architecture that impact the perceptions of agents.

This work contributes to the HCI community by presenting an in-depth analysis on the affect of conversation architecture on perceptions of conversational agents. We created a taxonomy for perception aspects and conversation architecture elements, providing clear and consistent terms that can be used in future research. Also, our analysis extends existing reviews on UX research of CAs \cite{clark2019state}\cite{rapp2021human}\cite{zheng2022ux} by providing a synthesized framework outlining the relationships between conversation architecture and perceptions of agents. With the help of this framework, we are taking the first step to understand how to design conversational agents to solicit specific social reactions from users.

In the remainder of the paper, we first outline our data collection and data analysis process, as well as the characteristics of the papers in our reviewed corpus. Following that, we describe in detail the taxonomy for the perception of agents, the taxonomy for conversation architecture element, as well as the framework on the relationships between perceptions of agents and conversation architecture. Finally, we discuss research challenges and opportunities as well as ethical considerations for designing conversational architecture to orchestrate perceptions of conversational agents.

%\section{Literature Review}

%\cite{finch2020towards} Towards unified dialogue system evaluation: A comprehensive analysis of current evaluation protocols

@@

Following PRISMA guidelines \cite{prisma}, we reviewed and selected literature exploring the relationship between the specifics of conversation architecture elements and perceptions of CAs. Searches were carried out in the ACM Digital Library between January 1 and January 15, 2023. Through our analysis, we didn't find any common terms used to refer to either conversation architecture elements, or aspects of perceptions. As such, we kept our search term to capture articles related to conversational agents. Based on search terms used in previously published literature reviews \cite{clark2019state}\cite{rapp2021human}, the following keywords are used to search for publications related to conversational user interfaces:
\newline

@@

\caption{Perception measures}

>>>>

Conversational agents (CAs), such as Amazon Alexa or Google Assistant, are designed to interact with humans using natural language through voice or text modalities. Since these agents interact with users using natural language, conversation architecture — which refers to what a CA says and how it says it — affects how these agents interact with users. The effectiveness of these interactions with CAs are strongly dependent on users' perceptions of agents \cite{knijnenburg2016inferring}\cite{seeger2021chatbots}. According to previous research, there is a correlation between CAs' conversation architecture and users' perception of these agents.

While there are various studies on how different conversation architecture elements affects agents' perceptions, there is a paucity of knowledge on the overall coverage and effects of these relationships explored in literature. For conversation architecture elements, The use of affective language is commonly explored, with studies finding that conversational agents are perceived as more socially present and emotionally intelligent if they use sentiment-adaptive response based on user's utterances \cite{diederich2019emulating}\cite{yang2017perceived}. Another commonly explored element is the use of prosody in voice-based agents, such as the use of express prosody contributing to higher perceived enjoyment and intimacy with the agent \cite{kim2020can}. However, there is a lack of research to synthesize these findings into an overall framework. One of the possible issues is the lack of clear and consistent definitions on the aspects of user's perceptions of agent and conversation architectural elements.

Currently there is no systematic understanding on the elements of conversation architecture that are relevant to perceptions of agents. Some papers have started to examine the speech specifics that are being explored in literature, such as Clark et al. \cite{clark2019state} capturing different elements of system speech production for voice-based agents and categorizing them based on content, style, and other topics. There are also some investigations of conversation architecture elements within the broader field of studying cues that trigger a social reaction of the user towards the CA. Feine et al's \cite{feine2019taxonomy} paper laid the foundation to create a consistent taxonomy for these social cues. However, there is no comprehensive research specifically for conversation architecture elements that are relevant to user perceptions.

Related to user perceptions, there is also no systematic understanding on the perceptions of agents affected by speech variations. Some research discussed the perceptions of agents as part of the broader scope of user experiences of CAs. These publications have gathered UX dimensions used to evaluate conversational agents, which include the perceptions of agents (e.g. empathy, humanness) alongside other assessments (e.g. physiological data) \cite{clark2019state}\cite{finch2020towards}. Some even categorized the metrics based on the type of method used, including technical measures like system error rate and perception measures like perceived social presence \cite{zheng2022ux}. However, these studies are assessing the general field of UX measurements and not specifically on the perceptions of agents, and do not analyze the effects of conversation architecture on these perceptions.

These research gaps discussed above: the paucity of knowledge on synthesizing the effect of speech specifics on perceptions of CAs as well as the lack of systematic understanding on the elements of conversation architecture and perceptions of agents motivated us to the goal of how to design conversation architecture in order to orchestrate perceptions of conversational agents. First, we want to understand \textbf{(RQ1) what perception aspects of agents are being explored in relation to the effects of conversation architecture elements}. While there are various architecture elements of conversation architecture used in a CA's utterances, we don't have a holistic understanding of them. This leads to our second research question: \textbf{(RQ2) what elements of conversation architecture are relevant to the perceptions of agents?} Lastly, we want to understand the effect of conversation architecture on perceptions of agent, specifically: \textbf{(RQ3) what are the relationships between conversation architecture elements and the perceptions of CAs?}

To answer these questions, we performed a comprehensive review on existing literature from 2010 to 2022 and synthesized the effect of conversation architecture elements on the perceptions of CAs based on 57 relevant papers (Figure \ref{fig:heatmap-impact}). We also grouped the perceptions of agents affected by conversation architecture into four categories and eleven aspects (Table \ref{tab:perceptions}), and grouped the speech variations related to perceptions of agents into four categories and eleven elements (Table \ref{tab:cues}). %Lastly, we coded the explored connections between conversation architecture and perceptions and synthesized their relationships into a framework. Through our analysis, we also found that there are inconsistencies in the assessment of perceptions across studies, as well as influencing factors other than conversation architecture that impact the perceptions of agents. 
This work contributes to the HCI community by presenting a framework that demonstrates the effect of conversation architecture on perceptions of conversational agents, as well as taxonomies for perception aspects and conversation architecture elements providing clear and consistent terms that can be used in future research. Together, these contributions move us one step closer to designing CAs that effectively interact with users. 
%Also, our analysis extends existing reviews on UX research of CAs \cite{clark2019state}\cite{rapp2021human}\cite{zheng2022ux} by providing a synthesized framework outlining the relationships between conversation architecture and perceptions of agents. With the help of this framework, we are taking the first step to understand how to design conversational agents to solicit specific social reactions from users.

In the remainder of the paper, we first outline related background work, our literature review process, as well as the characteristics of the papers in our reviewed corpus. Following that, we describe in detail the taxonomy for the perception of agents, the taxonomy for conversation architecture element, as well as the framework on the relationships between perceptions of agents and conversation architecture. Finally, we discuss research challenges and opportunities as well as ethical considerations for designing conversational architecture to orchestrate perceptions of conversational agents.

\section{Literature Review}

* human-human vs. human-AI

* anthropomorphism
%Anthropomorphism is defined as the attribution of traits that we typically associate with being distinctly human to nonhuman entities \cite{waytz2010sees}. For example, users tend to personify conversational agents like Amazon Alexa by using person pronouns instead of object pronouns in online reviews \cite{purington2017alexa}.  Research shows that anthropomorphic design is beneficial to establishing and maintaining trust between users and conversational agents \cite{seeger2021chatbots}. Also, speakers tend to align their lexical choices with conversational agents, similar to human-human conversations \cite{cowan2015does}. Anthropomorphism has been shown to ease user interactions, e.g. making agents more approachable, engaging, and trustworthy when they exhibit human characteristics \cite{qiu2009evaluating}. These processes are of particular interest for human-agent communication because they form the basis for user expectations and predispositions regarding agents' behavior, reliability of the information, etc. \cite{kuzminykh2020genie}.

* how to synthesize

* user experience evaluations of conversational agents
%Other papers categorized CA assessments into themes. For example, Rapp et al. \cite{rapp2021human} generated themes related to how users' experience with chatbots such as experiencing the chatbot including assessments on expectation, perception, satisfaction etc.  Most recently, Zheng et al's literature review on the UX research of CAs \cite{zheng2022ux} categorized the metrics based on the type of method used, including technical measures like system error rate and perception measures like perceived social presence. However, existing studies use different terminologies and categories across aspects of perception. As an illustration, it is unclear if the theme of experiencing the chatbot from Rapp et al. \cite{rapp2021human} contains similar aspects to the perception towards CA in Zheng et al \cite{zheng2022ux}'s analysis.
%\cite{finch2020towards} Towards unified dialogue system evaluation: A comprehensive analysis of current evaluation protocols

@@

Following PRISMA guidelines \cite{prisma}, we reviewed and selected literature exploring the relationship between the specifics of conversation architecture elements and perceptions of CAs. Searches were carried out in the ACM Digital Library between January 1 and January 15, 2023. Through our analysis, we didn't find any common terms used to refer to either conversation architecture elements, or aspects of perceptions. As such, we generalized our search term to capture articles related to conversational agents. Based on search terms used in previously published literature reviews \cite{clark2019state}\cite{rapp2021human}, the following keywords are used to search for publications related to conversational user interfaces:
\newline

@@

\caption{Perception aspects}
